\vspace{-3em} \textbf{ Mapping stellar surfaces\\ II: An interpretable Gaussian process model for light curves }

### 2021-01-28 paper2/ms.tex
addressed chedghes' comments; add hogg as author
better intro for paper2

% Author list
\author[0000-0002-0296-3826]{Rodrigo Luger}\altaffiliation{Flatiron Fellow}
\email{rluger@flatironinstitute.org}
\affil{Center~for~Computational~Astrophysics,~Flatiron~Institute,~New~York,~NY}
\affil{Virtual~Planetary~Laboratory, University~of~Washington, Seattle, WA}
%
\author[0000-0002-9328-5652]{Daniel Foreman-Mackey}
\affil{Center~for~Computational~Astrophysics,~Flatiron~Institute,~New~York,~NY}
%
\author[0000-0002-3385-8391]{Christina Hedges}
\affil{Bay~Area~Environmental~Research~Institute,~P.O.~Box~25,~Moffett~Field,~CA~94035,~USA}
\affil{NASA~Ames~Research~Center,~Moffett~Field,~CA}
%

@@

Over the past decades, Gaussian processes (GPs) have gained traction as
a leading tool for modeling correlated signals in astronomical time series datasets,
especially photometric time series
\citep[e.g.,][]{RasmussenWilliams2005,BrewerStello2009,Luger2016,ForemanMackey2017,Angus2018}
and radial velocities (ADD CITATIONS).
GPs are popular models for these applications because they allow marginalization
over a stochastic noise process specified only by its autocorrelation structure
or kernel.
This is relatively straightforward to interpret and in some specific cases it
can be related to physically motivated models (CITE SOME THINGS).
Furthermore, there are several popular open source implementations that allow
efficient evaluation of these models, and these have been widely demonstrated
to be useful effective models for the time series when the stochastic
variability is primarily a nuisance.

@@

If, however, one hopes to make inferences about the stellar surface itself,
GP models, as they are typically applied, are not so useful.
\textbf{While it is straightforward to derive
    posterior constraints on the hyperparameters of an effective GP model for
    observations of a star, it is not clear what those constraints actually
    tell us about the stellar surface.}
Specifically, in all but a few restricted cases, there is no first principles
relationship between the descriptive parameters of a typical GP model
(see section WHATEVER below)
and the physical properties of the stellar surface surface that is being observed.
For instance, it may be tempting to interpret the amplitude of the variability
as some measure of the spot contrast or the total number of spots,
or the timescale of the variations as the
spot lifetime, but there are no guarantees these interpretations will
hold in general. After all, the choice of kernel is quite often \emph{ad hoc},
providing an \emph{effective}---as opposed to \emph{interpretable}---description
of the physics.
%
There are two important exceptions to this: asteroseismic studies, in which the
the GP hyperparameters can offer direct insight into
the behavior of complex pulsation modes and thus physical properties of the
stellar interior \citep[e.g.,][]{BrewerStello2009,ForemanMackey2017}; and
stellar rotation period studies, in which the period hyperparameter $P$ can usually
be associated with the rotation period of the star \citep[e.g.,][]{Angus2018}.%
\footnote{An exception to this is in the presence of strong differential
    rotation, in which case many periods may be present in the data, or
    when spots evolve coherently, which can also introduce weak periodicities
    in the light curve.}
%
For spot-induced variability, on the other hand,
GPs are usually used
when the variability itself is a nuisance parameter. For example, if the
goal is to constrain the properties of a transiting exoplanet or to
search for a planetary signal in a radial velocity dataset, a GP might be
used to remove (or, better yet, to marginalize over) the stellar variability
\citep[e.g.,][]{Haywood2014,Rajpaul2015,Luger2017b}.
In this case, the physics behind the variability is irrelevant, so an
effective model of this sort is generally sufficient.

When the goal is to understand
the stellar surface itself, however, GPs have been less useful to date.
While recent papers have proposed several custom
GP kernels that excel at modeling rotational variability, such as
the exponential sine squared kernel \citep[Equation~\ref{eq:expsinsq};][]{Wang2012},
the quasi-periodic kernel \citep{Angus2018}, and
the \celerite \textsf{RotationTerm} \citep{ForemanMackey2017},
the connection between the kernel hyperparameters and the physical spot
properties is often qualitative. Recently, \citet{Perger2020} proposed
an empirically-motivated quasi-periodic cosine kernel that can be used
to infer spot lifetimes in the absence of differential rotation. However,
an explicit connection between the GP hyperparameters and the size and
spatial distribution of spots has not been explored in the literature.

Instead, a common approach to inferring starspot properties from
photometric datasets is
to forward model the stellar surface and its effect on the light curve, then
seek to constrain the surface properties by solving the inverse problem.
We discussed this approach in \citet{PaperI} \citepalias[hereafter][]{PaperI},
where we argued a unique solution to the surface map of
the star from a rotational light curve is not possible without the use of
aggressive (and often \emph{ad hoc}) priors. The degeneracies at play
make it effectively impossible for one to know the exact configuration
of starspots and other features on the surface of a star from its rotational
light curve alone.

However, it is hardly ever the case that this is actually our end goal.
After all, physics can be used to predict properties of stellar surfaces at a fairly
high level: i.e., typical spot sizes, active spot latitudes, or approximate
timescales on which spots evolve
\citep[e.g.,][]{Schuessler1996,Solanki2006,Cantiello2019}.
We are hardly ever interested in the
\emph{particular} properties of a \emph{particular} spot, as we wouldn't really
know what to do with that information! Instead, we often treat
(whether explicitly or not)
the properties of a starspot as a draw from some parent distribution
controlling (say) the average and spread in the radii of the spots.
The parameters controling this distribution are the ones that we can
predict with physics; they are therefore also the ones we are usually
interested in.

@@

\subsection{Gaussian processes in general}

Despite whatever mystique the words ``Gaussian process'' may evoke, a GP
is nothing but a Gaussian distribution in many (formally infinite)
dimensions. Specifically, it is a Gaussian distribution over
\emph{functions} spanning a continuous domain (in our case, the time domain).
Similar to a multivariate Gaussian, which is described by
a $(K \times 1)$ vector $\pmb{\mu}$ characterizing
the mean of the process and a $(K \times K)$
matrix $\pmb{\Sigma}$ characterizing its covariance,
a GP is fully specified by a mean
function $m(t)$ and a kernel function $k(t, t')$.
To say that a random vector-valued variable $\mathbbb{f}$
defined on a $(K \times 1)$ time array $\mathbf{t}$
is ``distributed as a GP'' means that we may write
%
\begin{align}
    \label{eq:fnormal}
    \mathbbb{f} \sim \mathcal{N}\left( \pmb{\mu}, \pmb{\Sigma} \right)
    \quad,
\end{align}
%
where the elements of the mean and covariance are given by
$\mu_i = m(t_i)$ and $\Sigma_{i,j} = k(t_i, t_j)$, respectively.%
\footnote{%
    In this paper, we will use blackboard font (i.e., $\mathbbb{f}$) to
    denote random variables and serif font (i.e., $\mathbf{f}$) to denote
    particular realizations of those variables.
    See Appendix~\ref{sec:notation} for a detailed explanation of our notation.
}
Because of this relationship to multivariate Gaussians,
GPs are easy to sample from.%
\footnote{Given a 1-d array \texttt{mean} and a 2-d array \texttt{cov} in \Python,
    sampling from the corresponding GP (if it exists)
    can be done in a single line of code by calling
    \texttt{numpy.random.multivariate\_normal(mean, cov)}.}
But the real showstopper is the application of GPs to inference problems.
Multivariate Gaussian distributions have a closed-form (marginal) likelihood
function, so it is easy to compute the probability
of one's data conditioned on a given value of $\pmb{\mu}$ and $\pmb{\Sigma}$
(i.e., the ``likelihood''; see Equation~\ref{eq:log-like} below).
This can in turn be maximized
to infer the optimal values of the model parameters
or used in a
numerical sampling scheme to compute the probability of those parameters
given the data (i.e., the ``posterior'').
Thanks to modern computer architectures, linear algebra packages, and
GP algorithms,
evaluating the GP likelihood may typically be done in a fraction of a second
for a reasonably-sized dataset (i.e., $K \lesssim 10^4$ datapoints).

Another big advantage of GPs is their flexibility. GPs are often dubbed
a class of ``non-parametric'' models, given that nowhere in the specification
of the GP is there an explicit functional form for $\mathbbb{f}$. Rather, a GP
is a stochastic
process whose draws can in principle take on \emph{any} functional form,
subject, however, to certain smoothness and correlation criteria
of tunable strictness
that are fully
encoded in the covariance $\pmb{\Sigma}$.
%
In many applications, particularly when modeling stellar light curves,
it is customary to restrict the problem by
assuming that the process is \emph{stationary}, such that we may write
%
\begin{align}
    \label{eq:kernel}
    \Sigma_{i,j} & = k(t_i, t_j)
    \nonumber                                    \\
                 & = k(\left| t_i - t_j \right|)
    \nonumber                                    \\
                 & \equiv k(\Delta t)
    \quad.
\end{align}
%
A stationary process is one that is independent of phase (or, in this case,
the actual value of the time $t$); rather, it depends only on the \emph{difference}
between the phases of two data points. The kernel of a stationary process is
therefore a one-dimensional function, typically chosen from a set of
standard functions with desirable smoothness and spectral properties.
Among a vast number of kernels are a few that are commonly used
in modeling stellar variability: the exponential squared kernel,
%
\begin{align}
    \label{eq:expsq}
    k_\mathrm{E^2}(\Delta t)           & = \sigma^2 \exp\left(-\frac{\Delta t^2}{2\tau}\right)
    \quad,                                                                                                                                           \\
    \intertext{the Mat\'ern-3/2 kernel,}
    \label{eq:mat32}
    k_\mathrm{M^\frac{3}{2}}(\Delta t) & = \sigma^2 \left( 1 + \frac{\sqrt{3}\Delta t}{\tau} \right) \exp\left(-\frac{\sqrt{3}\Delta t}{\tau}\right)
    \quad,                                                                                                                                           \\
    \intertext{the exponential sine squared kernel,}
    \label{eq:expsinsq}
    k_\mathrm{S^2}(\Delta t)           & = \sigma^2 \exp\left[-\Gamma\sin^2\left( \frac{\pi \Delta t}{P} \right)\right]
    \quad,                                                                                                                                           \\
    \intertext{the cosine kernel,}
    k_\mathrm{C}(\Delta t)             & = \sigma^2 \cos\left( \frac{2\pi \Delta t}{P} \right)
    \quad,
\end{align}
%
or any linear combination and/or product of these. In the expressions above,
$\sigma$, $\tau$, $\Gamma$, and $P$ are
the \emph{hyperparameters} of the kernels.
Specifically, $\sigma^2$ is the variance (the value
of the kernel when $t_i = t_j$), $\tau$ is a typical timescale over which
data are correlated, $\Gamma$ is a dimensionless scale parameter, and $P$ is
a period.
Once a kernel has been chosen, these scalar quantities fully
specify the covariance of the Gaussian process. GP-based inference
entails optimizing or sampling over these few quantities, which in many
cases can be done easily and efficiently.

>>>>

% Author list
\author[0000-0002-0296-3826]{Rodrigo Luger}\altaffiliation{Flatiron Fellow}
\email{rluger@flatironinstitute.org}
\affil{Center~for~Computational~Astrophysics,~Flatiron~Institute,~New~York,~NY}
\affil{Virtual~Planetary~Laboratory, University~of~Washington, Seattle, WA}
%
\author[0000-0002-9328-5652]{Daniel Foreman-Mackey}
\affil{Center~for~Computational~Astrophysics,~Flatiron~Institute,~New~York,~NY}
%
\author[0000-0002-3385-8391]{Christina Hedges}
\affil{Bay~Area~Environmental~Research~Institute,~Moffett~Field,~CA}
\affil{NASA~Ames~Research~Center,~Moffett~Field,~CA}
%

@@

Over the past two decades, Gaussian processes (GPs) have gained traction as
a leading tool for modeling correlated signals in astronomical datasets
\citep[][]{RasmussenWilliams2005}.
In particular, GPs are commonly used to model
stellar variability in photometric time series
\citep[e.g.,][]{BrewerStello2009,Aigrain2016,Luger2016,ForemanMackey2017,Angus2018}
and radial velocity measurements \citep[e.g.,][]{Rajpaul2015,Jones2017,Perger2020}.
GPs are popular models for these applications because they allow marginalization
over a stochastic noise process specified only by a kernel describing its autocorrelation structure.
%
There are several popular open source implementations that allow
efficient evaluation of GPs, and these have been widely demonstrated
to be useful effective models for the time series when the stochastic
variability due to the star is primarily a nuisance
\citep[e.g.,][]{Ambikasaran2015,ForemanMackey2017,Gilbertson2020}.

@@

A major source of stellar variability in both light curves and radial velocity
datasets is the modulation induced by surface features like starspots
rotating in and out of view. While GPs excel at marginalizing over stellar rotational
variability, they have been less useful when the goal is to make
inferences about the actual source of this variability, such as the
properties of starspots and the magnetic processes that generate them.
\textbf{While it is straightforward to derive
    posterior constraints on the hyperparameters of an effective GP model for
    observations of a star, it is not clear what those constraints actually
    tell us about the stellar surface.}
Specifically, in all but a few restricted cases, there is no \emph{first principles}
relationship between the descriptive parameters of a typical GP model
(see \S\ref{sec:gp-intro})
and the physical properties of the stellar surface that is being observed.
%
For instance, it may be tempting to interpret the GP amplitude hyperparameter
as some measure of the spot contrast or the total number of spots,
or the GP timescale hyperparameter as the
spot lifetime, but there are no guarantees these interpretations will
hold in general. After all, the choice of kernel is quite often \emph{ad hoc},
providing an \emph{effective}---as opposed to \emph{interpretable}---description
of the physics.
%
There are two important exceptions to this: asteroseismic studies, in which the
the GP hyperparameters can offer direct insight into
the behavior of complex pulsation modes and thus physical properties of the
stellar interior \citep[e.g.,][]{BrewerStello2009,ForemanMackey2017}; and
stellar rotation period studies, in which the period hyperparameter $P$ can usually
be associated with the rotation period of the star \citep[e.g.,][]{Angus2018}.%
\footnote{An exception to this is in the presence of strong differential
    rotation, in which case many periods may be present in the data, or
    when spots evolve coherently, which can also introduce weak periodicities
    in the light curve.}
%
For spot-induced variability, on the other hand,
GPs are usually used
when the variability itself is a nuisance parameter. For example, if the
goal is to constrain the properties of a transiting exoplanet or to
search for a planetary signal in a radial velocity dataset, a GP might be
used to remove (or, better yet, to marginalize over) the stellar variability
\citep[e.g.,][]{Haywood2014,Rajpaul2015,Luger2017b}.
In this case, the physics behind the variability is irrelevant, so an
effective model of this sort is generally sufficient.

Understanding the properties of starspots, faculae, plages, and other
features on the surfaces of stars is crucial to \xxx{UNDERSTANDING MAGNETIC
    FIELDS, STELLAR INTERIORS, MITIGATING EXOPLANET SYSTEMATICS} \citep{Rackham2018}.

% Recent papers have proposed several custom
% GP kernels that excel at modeling rotational variability, such as
% the exponential sine squared kernel \citep[Equation~\ref{eq:expsinsq};][]{Wang2012},
% the quasi-periodic kernel \citep{Angus2018}, and
% the \celerite \textsf{RotationTerm} \citep{ForemanMackey2017}.
% However, the connection between the kernel hyperparameters and the physical spot
% properties is often qualitative. Recently, \citet{Perger2020} proposed
% an empirically-motivated quasi-periodic cosine kernel that can be used
% to infer spot lifetimes in the absence of differential rotation. However,
% an explicit connection between the GP hyperparameters and the size and
% spatial distribution of spots has not been explored in the literature.

When the goal is to learn about the stellar surface, the common approach
in the literature
has been to explicitly \emph{forward model} the stellar surface and its effect on the light curve, then
seek to constrain the surface properties by solving the inverse problem.
We discussed this approach in \citet{PaperI} \citepalias[hereafter][]{PaperI},
where we argued a unique solution to the surface map of
the star from a rotational light curve is not possible without the use of
aggressive (and often \emph{ad hoc}) priors. The degeneracies at play
make it effectively impossible for one to know the exact configuration
of starspots and other features on the surface of a star from its rotational
light curve alone.

However, it is hardly ever the case that this is actually our end goal.
After all, physics can be used to predict properties of stellar surfaces at a fairly
high level: i.e., typical spot sizes, active spot latitudes, or approximate
timescales on which spots evolve
\citep[e.g.,][]{Schuessler1996,Solanki2006,Cantiello2019}.
We are hardly ever interested in the
\emph{particular} properties of a \emph{particular} spot, as we wouldn't really
know what to do with that information! Instead, we often treat
(whether explicitly or not)
the properties of a starspot as a draw from some parent distribution
controlling (say) the average and spread in the radii of the spots.
The parameters controlling this distribution are the ones that we can
predict with physics; they are therefore also the ones we are usually
interested in.

@@

\subsection{Brief overview of Gaussian processes}
\label{sec:gp-intro}
Despite whatever mystique the words ``Gaussian process'' may evoke, a GP
is nothing but a Gaussian distribution in many (formally infinite)
dimensions. Specifically, it is a Gaussian distribution over
\emph{functions} spanning a continuous domain (in our case, the time domain).
Similar to a multivariate Gaussian, which is described by
a $(K \times 1)$ vector $\pmb{\mu}$ characterizing
the mean of the process and a $(K \times K)$
matrix $\pmb{\Sigma}$ characterizing its covariance,
a GP is fully specified by a mean
function $m(t)$ and a kernel function $k(t, t')$.
To say that a random vector-valued variable $\mathbbb{f}$
defined on a $(K \times 1)$ time array $\mathbf{t}$
is ``distributed as a GP'' means that we may write
%
\begin{align}
    \label{eq:fnormal}
    \mathbbb{f} \sim \mathcal{N}\left( \pmb{\mu}, \pmb{\Sigma} \right)
    \quad,
\end{align}
%
where the elements of the mean and covariance are given by
$\mu_i = m(t_i)$ and $\Sigma_{i,j} = k(t_i, t_j)$, respectively.%
\footnote{%
    In this paper, we will use blackboard font (i.e., $\mathbbb{f}$) to
    denote random variables and serif font (i.e., $\mathbf{f}$) to denote
    particular realizations of those variables.
    See Appendix~\ref{sec:notation} for a detailed explanation of our notation.
}
Because of this relationship to multivariate Gaussians,
GPs are easy to sample from.%
\footnote{Given a 1-d array \texttt{mean} and a 2-d array \texttt{cov} in \Python,
    sampling from the corresponding GP (if it exists)
    can be done in a single line of code by calling
    \texttt{numpy.random.multivariate\_normal(mean, cov)}.}
But the real showstopper is the application of GPs to inference problems.
Multivariate Gaussian distributions have a closed-form (marginal) likelihood
function, so it is easy to compute the probability
of one's data conditioned on a given value of $\pmb{\mu}$ and $\pmb{\Sigma}$
(i.e., the ``likelihood''; see Equation~\ref{eq:log-like} below).
This can in turn be maximized
to infer the optimal values of the model parameters
or used in a
numerical sampling scheme to compute the probability of those parameters
given the data (i.e., the ``posterior'').
Thanks to modern computer architectures, linear algebra packages, and
GP algorithms,
evaluating the GP likelihood may typically be done in a fraction of a second
for a reasonably-sized dataset (i.e., $K \lesssim 10^4$ datapoints).

Another big advantage of GPs is their flexibility. GPs are often dubbed
a class of ``non-parametric'' models, given that nowhere in the specification
of the GP is there an explicit functional form for $\mathbbb{f}$. Rather, a GP
is a stochastic
process whose draws can in principle take on \emph{any} functional form,
subject, however, to certain smoothness and correlation criteria
of tunable strictness
that are fully
encoded in the covariance $\pmb{\Sigma}$.
%
In many applications, particularly when modeling stellar light curves,
it is customary to restrict the problem by
assuming that the process is \emph{stationary}, such that we may write
%
\begin{align}
    \label{eq:kernel}
    \Sigma_{i,j} & = k(t_i, t_j)
    \nonumber                                    \\
                 & = k(\left| t_i - t_j \right|)
    \nonumber                                    \\
                 & \equiv k(\Delta t)
    \quad.
\end{align}
%
A stationary process is one that is independent of phase (or, in this case,
the actual value of the time $t$); rather, it depends only on the \emph{difference}
between the phases of two data points. The kernel of a stationary process is
therefore a one-dimensional function, typically chosen from a set of
standard functions with desirable smoothness and spectral properties.
\xxx{REMOVE THESE EXAMPLES; NOT NEEDED.}
Among a vast number of kernels are a few that are commonly used
in modeling stellar variability: the exponential squared kernel,
%
\begin{align}
    \label{eq:expsq}
    k_\mathrm{E^2}(\Delta t)           & = \sigma^2 \exp\left(-\frac{\Delta t^2}{2\tau}\right)
    \quad,                                                                                                                                           \\
    \intertext{the Mat\'ern-3/2 kernel,}
    \label{eq:mat32}
    k_\mathrm{M^\frac{3}{2}}(\Delta t) & = \sigma^2 \left( 1 + \frac{\sqrt{3}\Delta t}{\tau} \right) \exp\left(-\frac{\sqrt{3}\Delta t}{\tau}\right)
    \quad,                                                                                                                                           \\
    \intertext{the exponential sine squared kernel,}
    \label{eq:expsinsq}
    k_\mathrm{S^2}(\Delta t)           & = \sigma^2 \exp\left[-\Gamma\sin^2\left( \frac{\pi \Delta t}{P} \right)\right]
    \quad,                                                                                                                                           \\
    \intertext{the cosine kernel,}
    k_\mathrm{C}(\Delta t)             & = \sigma^2 \cos\left( \frac{2\pi \Delta t}{P} \right)
    \quad,
\end{align}
%
or any linear combination and/or product of these. In the expressions above,
$\sigma$, $\tau$, $\Gamma$, and $P$ are
the \emph{hyperparameters} of the kernels.
Specifically, $\sigma^2$ is the variance (the value
of the kernel when $t_i = t_j$), $\tau$ is a typical timescale over which
data are correlated, $\Gamma$ is a dimensionless scale parameter, and $P$ is
a period.
Once a kernel has been chosen, these scalar quantities fully
specify the covariance of the Gaussian process. GP-based inference
entails optimizing or sampling over these few quantities, which in many
cases can be done easily and efficiently.
\xxx{BETTER TRANSITION HERE.}
